---
title: "Portfolio Weights before and after RMT filtering March 14 2025"
author: "Sulalitha Bowala"
date: "2025-03-14"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Here’s an implementation of RMT-based covariance estimation in R, using eigenvalue filtering to improve financial time series analysis.


Steps:
1. Install & Load Required Libraries
2. Generate Synthetic Financial Data: We simulate N assets over T time periods with an underlying true covariance structure.
3. Compute Eigenvalues & Eigenvectors
4. Marchenko-Pastur Theoretical Bounds: According to RMT, random covariance matrices follow the Marchenko-Pastur (MP) law.
The theoretical bounds for eigenvalues are:
$$\lambda_{\text{min}, \text{max}} = (1 \pm \sqrt{Q})^2$$
5. Filter Eigenvalues & Reconstruct Covariance Matrix
•	Keep only the informative eigenvalues (those outside the MP bulk).
•	Replace noisy eigenvalues with their mean.
6. Plot Eigenvalue Distribution: This helps visualize the Marchenko-Pastur bulk and identify informative eigenvalues.
7. Compare Portfolio Optimization Performance: We test how the denoised covariance matrix improves portfolio allocation stability.
Compute Portfolio Weights Using Markowitz Optimization
8. Results & Interpretation
•	Eigenvalue filtering removes noisy correlations, stabilizing the covariance matrix.
•	Denoised covariance matrix improves portfolio optimization, reducing overfitting to noise.
•	More robust asset allocation, as seen in the smoother weight distribution.


```{r}
# Load required libraries
library(ggplot2)
library(MASS)
library(quadprog)
library(Matrix)  # For nearPD function

# Set random seed for reproducibility
set.seed(42)

# Parameters
N <- 100    # Number of assets
T <- 300    # Number of time periods
Q <- N / T  # Aspect ratio

# Generate true covariance matrix
true_cov <- matrix(rnorm(N * N), nrow = N, ncol = N)
true_cov <- (true_cov + t(true_cov)) / 2  # Make it symmetric
#diag(true_cov) <- abs(diag(true_cov)) + 0.5  # Add positive values to diagonal
# Why adding 0.5 to all diagonals
# Ensure positive definiteness
true_cov <- as.matrix(nearPD(true_cov)$mat)

# Generate returns
returns <- mvrnorm(n = T, mu = rep(0, N), Sigma = true_cov)

# Compute empirical covariance matrix
C <- cov(returns)

# Perform eigendecomposition
eigen_decomp <- eigen(C)
eigvals <- eigen_decomp$values
eigvecs <- eigen_decomp$vectors

# Calculate Marchenko-Pastur bounds
lambda_min <- (1 - sqrt(Q))^2
lambda_max <- (1 + sqrt(Q))^2

# Identify informative eigenvalues (there aren't any id we don't add 0.5?)
informative <- eigvals > lambda_max

# Calculate mean of the bulk eigenvalues
bulk_mean <- mean(eigvals[!informative])

# RMT filtering function
filter_rmt <- function(C) {
  # Perform eigendecomposition
  eig <- eigen(C)
  eigvals <- eig$values
  eigvecs <- eig$vectors
  
  # Calculate MP bounds
  Q <- ncol(C) / nrow(C)
  lambda_min <- (1 - sqrt(Q))^2
  lambda_max <- (1 + sqrt(Q))^2
  
  # Identify informative eigenvalues
  informative <- eigvals > lambda_max
  bulk_mean <- mean(eigvals[!informative])
  
  # Reconstruct filtered matrix
  C_filtered <- eigvecs[, informative] %*% diag(eigvals[informative]) %*% t(eigvecs[, informative])
  C_filtered <- C_filtered + bulk_mean * diag(nrow(C))
  
  return(C_filtered)
}

# Construct filtered covariance matrix
C_filtered <- filter_rmt(C)

# MSE function for comparison
mse <- function(A, B) {
  return(mean((A - B)^2, na.rm = TRUE))
}

# Calculate MSE between true and estimated covariances
mse_original <- mse(C, true_cov)
mse_filtered <- mse(C_filtered, true_cov)

# Print MSE results
print(paste("MSE (Original):", mse_original))
print(paste("MSE (RMT Filtered):", mse_filtered))

# Visualization of eigenvalue distribution
df <- data.frame(Eigenvalues = eigvals)
p <- ggplot(df, aes(x = Eigenvalues)) +
  geom_histogram(binwidth = 0.1, fill = "blue", alpha = 0.6, color = "black") +
  geom_vline(xintercept = lambda_min, color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = lambda_max, color = "red", linetype = "dashed", size = 1) +
  ggtitle("Eigenvalue Distribution vs. Marchenko-Pastur Boundaries") +
  xlab("Eigenvalue") + ylab("Frequency") +
  theme_minimal()

# Display the plot
print(p)

# Portfolio optimization comparison
# Compute inverse of covariance matrices
inv_C <- solve(C)
inv_C_filtered <- solve(C_filtered)

# Assume equal expected returns
mu <- rep(1, N)

# Compute portfolio weights
w_opt <- inv_C %*% mu / sum(inv_C %*% mu)
w_opt_filtered <- inv_C_filtered %*% mu / sum(inv_C_filtered %*% mu)

# Plot portfolio weights comparison
plot(w_opt, type = "l", col = "red", lwd = 2,
     ylim = c(min(w_opt, w_opt_filtered), max(w_opt, w_opt_filtered)),
     main = "Portfolio Weights Before and After RMT Filtering",
     ylab = "Weight", xlab = "Asset Index")
lines(w_opt_filtered, col = "blue", lwd = 2)
legend("topright", legend = c("Original", "RMT-Filtered"),
       col = c("red", "blue"), lwd = 2)
```


Conclusion:

•	RMT-based filtering is an effective method for improving covariance estimation in financial data.
•	Applications include portfolio optimization, risk management, and factor modeling.
•	Future work: Extend to dynamic covariance estimation using rolling windows or deep learning.


